A full approximation scheme multilevel method \\ for nonlinear variational inequalities

### 2023-07-20 paper/fascd.tex
FAScd -> FASCD

\begin{abstract}
We present the \emph{full approximation scheme constraint decomposition} (FAScd) multilevel method for solving variational inequalities (VIs).  FAScd is a common extension both of the full approximation scheme (FAS) multigrid technique for nonlinear partial differential equations, due to A.~Brandt, and the constraint decomposition (CD) method introduced by X.-C.~Tai for VIs arising in optimization.  We extend the CD idea by exploiting the telescoping nature of certain function space subset decompositions arising from multilevel mesh hierarchies.  We demonstrate that FAScd V-cycles exhibit excellent mesh-independent convergence rates and that FAScd F-cycles (full multigrid) are optimal solvers, when used in combination with reduced-space active set smoothers with work proportional to the number of unknowns on a given mesh level.  The example problems include unilateral and bilateral VI problems, with differential operators which are symmetric linear, nonsymmetric linear, and nonlinear.  Our examples demonstrate very good convergence even on problems where other multilevel VI solvers struggle.
\end{abstract}

@@

The constraint decomposition (CD) methods of Tai \cite{Tai2003} are designed to solve coercive variational inequality (VI) problems arising as the Karush--Kuhn--Tucker optimality conditions for the minimization of a convex functional over a convex set.  These methods were shown to converge for multilevel and domain decompositions in a finite element (FE) context; in the multilevel case, the CD method has almost grid-independent error bounds, and near-optimal complexity, for elliptic, linear obstacle problems \cite[Subsection 5.4]{Tai2003}, \cite[Theorem 4.6 and Algorithm 4.7]{GraeserKornhuber2009}.
In this paper we extend Tai's multilevel CD method by replacing the coarser-level corrections with a full approximation scheme (FAS; \cite{Brandt1977,Bruneetal2015}) approach.  Specifically, our \emph{full approximation scheme constraint decomposition} (FAScd) method extends the algorithms in \cite{GraeserKornhuber2009} and \cite{Tai2003} in the following five directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems. The needed computations require neither an objective function nor linearity of the residual operator.
\item We identify a ``telescoping'' subset property which can be exploited on the up-smoothing side of a V-cycle for greater performance.
\item Box constraints, namely upper and lower obstacles, are permitted.
\item We implement V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (full multigrid; Algorithm \ref{alg:fascd-fcycle}).
\item We demonstrate that F-cycle FAScd is an optimal solver for the different VI examples in Section \ref{sec:results}, when used in conjunction with reduced-space active set methods as smoothers. The number of Newton and Krylov iterations per step are fixed, so that the work of the smoother is proportional to the number of unknowns on a given level.
\end{enumerate}

Prior multilevel algorithms for VIs have various substantial limitations compared to the FAScd framework.  The projected full approximation scheme (PFAS) algorithm of Brandt \& Cryer \cite{BrandtCryer1983} is apparently limited to linear complementarity problems and projected (truncated) Gauss--Seidel smoothers.  While FAScd reduces to general, nonlinear FAS for PDEs when the inequality constraints are removed (Appendix \ref{app:reductions}), PFAS reduces to linear multigrid.
The multilevel CD iteration of Tai \cite{Tai2003} is formulated in terms of an objective function, so as stated it only applies to optimization-type problems, and furthermore each of its coarser-level problems refers to the finest-level discretization.  The problematic latter property is corrected by reformulation using defect constraints and monotone injection operators (see \cite[Algorithm 4.7]{GraeserKornhuber2009} and Section \ref{sec:cdmultilevel} herein).  The monotone multigrid methods of Kornhuber and co-authors \cite{GraeserKornhuber2009,Kornhuber1994} appear to be formulated only with Gauss--Seidel type smoothers, and applied only to optimization-type VI problems \cite[for example]{JouvetGraeser2013}.  Furthermore, monotone multigrid methods are more difficult to implement in FE libraries because modifications of the FE basis functions are needed to approximate the coarser-level coincidence sets.  No such invasive basis modifications are used either in Tai's original multilevel CD or in FAScd.  Our implementation of FAScd in the Firedrake library \cite{Rathgeberetal2016} requires no modifications of basis functions and yet exhibits excellent performance, including on a problem used to demonstrate the benefits of monotone multigrid (Example \ref{ex:results:classical}; compare \cite[problem 7.1.1]{GraeserKornhuber2009}).

In our experiments, FAScd V-cycles exhibit essentially the same mesh-independent convergence rates for VI problems that FAS multigrid V-cycles exhibit for the corresponding (unconstrained) PDE problems.  The F-cycle results in Section \ref{sec:results} show nearly-textbook multigrid efficiency, namely solutions within discretization error in work comparable to few smoother sweeps on the finest-level mesh \cite{BrandtLivne2011}.  Specifically, we see this performance for classical Laplacian obstacle problems, $p$-Laplacian obstacle problems, and nonnegativity-constrained advection-diffusion equations.  Furthermore, these results appear to be insensitive to the length or structure of the free boundary, i.e.~the geometric complexity of the coincidence set.

All iterates in FAScd are admissible, and thus the nonlinear operator in the VI needs only be defined for admissible states.  For example, we solve an ice sheet geometry problem (Examples \ref{ex:sia} and \ref{ex:results:sia}) in which the ice flow formulas are only meaningful for surface elevation iterates that do not penetrate the bedrock.  (More generally, fluid layers subject to a non-negative thickness constraint typically require the solution of an auxiliary PDE on a domain determined inside the VI residual evaluation \cite{Bueler2021conservation}, but this is not needed in the ice sheet problem here.)  Non-admissible methods, including semi-smooth methods \cite{BensonMunson2006}, would require unnatural modifications of the operator formula.  FAScd solves this low-regularity ice sheet problem, wherein the solution has unbounded gradient at the free boundary, by a mesh-independent number of F-cycle iterations (Example \ref{ex:results:sia}).

The paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone variational inequalities, and describe several motivating example problems.  Section \ref{sec:cd} reviews the basics of CDs and their associated iterations.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, and then Section \ref{sec:cdmultilevel} presents defect constraints generated via monotone injection operators \cite{GraeserKornhuber2009}.  Here we propose an apparently new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD and an associated iteration.  Section \ref{sec:vcycle} then states the FAScd V-cycle, with Section \ref{sec:implementation} addressing convergence criteria, F-cycles, and smoother construction.  Section \ref{sec:results} gives numerical results, from a parallel Firedrake \cite{Rathgeberetal2016} implementation (see Code Availability), for each example problem introduced in Section \ref{sec:vi}.  The work concludes with a discussion and outlook in Section \ref{sec:discussion}.

@@

Numerical results are shown in Section \ref{sec:results} for applications of FAScd Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} to these Examples.

@@

Associated to any CD are certain iterative methods \cite{Tai2003,Xu1992}.  Let $0\le \alpha \le 1$ be the \emph{damping parameter}.\footnote{FAScd Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} are presented without damping, and indeed it seems to be unnecessary.  However, the convergence of the general CD iterations appears to depend upon it.}  The following algorithms, which replace $u \in \cK$ with a new iterate $w\in\cK$ which should be closer to $u^\star \in \cK$ solving \eqref{eq:vi}, solve smaller VI problems over each subset $\cK_i$.

@@

\section{FAScd V-cycle} \label{sec:vcycle}

@@

\begin{figure}[ht]
\begin{center}
\input{tikz/vcycle.tex}
\end{center}
\caption{The FAScd V-cycle computes downward corrections $y_j \in \mathcal{D}^j$, but the upward corrections $z_j\in\mathcal{U}^j$ are in larger sets.}
\label{fig:fascdvcycle}
\end{figure}

@@

\begin{pseudofloat}[ht]
\begin{pseudo}
\pr{fascd-vcycle}(J,\ell^J,\underline{\gamma}^J,\overline{\gamma}^J;w^J)\text{:} \\+
    $\underline{\chi}^J, \, \overline{\chi}^J = \underline{\gamma}^J - w^J, \, \overline{\gamma}^J - w^J {\large \strut}$ \label{line:vcyclegenchifinest} \\
    for $j=J$ downto $j=1$ \\+
      $\underline{\chi}^{j-1}, \, \overline{\chi}^{j-1} = \maxR \underline{\chi}^j, \, \minR \overline{\chi}^j {\large \strut}$ \label{line:vcyclegenchi} \\
      $\underline{\phi}^j, \, \overline{\phi}^j = \underline{\chi}^j - P\underline{\chi}^{j-1}, \, \overline{\chi}^j - P\overline{\chi}^{j-1} {\large \strut}$ \label{line:vcyclegenphi} \\
      $y^j = 0$ \\
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j,\underline{\phi}^j,\overline{\phi}^j,w^j;y^j)$  \ct{smoothing of \eqref{eq:fe:downvi} in $\mathcal{D}^j$}\\
      $w^{j-1} = \iR(w^j + y^j)$ \label{line:vcyclerestrictsolution} \\
      $\ell^{j-1} = f^{j-1}(w^{j-1}) + R \left(\ell^j - f^j(w^j+y^j)\right)$ \label{line:vcyclerestrictell} \\-
    $z^0 = 0$ \\
    $\text{\pr{solve}}(\ell^0,\underline{\chi}^0,\overline{\chi}^0,w^0;z^0)$ \hspace{1.0cm} \ct{solving of \eqref{eq:fe:upvi} in $\mathcal{U}^0$} \\
    for $j=1$ to $j=J$ \\+
      $z^j = y^{j} + P z^{j-1}$ \label{line:vcycleupsmoothinitial} \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j,\underline{\chi}^j,\overline{\chi}^j,w^j;z^j)$  \ct{smoothing of \eqref{eq:fe:upvi} in $\mathcal{U}^j$} \\-
    return $w^J+z^J$
\end{pseudo}
\caption{The FAScd V-cycle is an iteration for solving VI problem \eqref{eq:fe:vi}.  Here $f^j$ denotes an FE discretization of $f$ in problem \eqref{eq:boxdirichletvi}.}
\label{alg:fascd}
\end{pseudofloat}

@@

Implementations of FAScd methods must choose a concrete convergence criterion.  In this section we also construct the F-cycle extension, and we describe the active set Newton method we will use as a smoother and coarse solver.

@@

\begin{pseudofloat}[h]
\begin{pseudo}
\pr{fascd-fcycle}(J,\ell^J,\underline{\gamma}^J,\overline{\gamma}^J)\text{:} \\+
    for $j=J$ downto $j=1$ \\+
        $\ell^{j-1} = R(\ell^j)$ \\
        $\underline{\gamma}^{j-1}, \, \overline{\gamma}^{j-1} = \iR \underline{\gamma}^{j}, \, \iR \overline{\gamma}^{j}$ {\large\strut} \\-
    choose $w^0 \in \mathcal{K}^0$ \label{line:fcyclecoarsestinitial} \\
    $\text{\pr{solve}}(\ell^0,\underline{\gamma}^0,\overline{\gamma}^0;w^0)$ \\
    for $j=1$ to $j=J-1$ \\+
        $w^j = \max\left\{\underline{\gamma}^{j},\min\{\overline{\gamma}^{j}, Pw^{j-1}\}\right\} \in \mathcal{K}^j$ \label{line:fcycleprolongtruncone} \\
        $\pr{fascd-vcycle}^{\text{\id{rampv}}}(j,\ell^j,\underline{\gamma}^j,\overline{\gamma}^j;w^j)$ \\-
    $w^J = \max\left\{\underline{\gamma}^{J},\min\{\overline{\gamma}^{J}, Pw^{J-1}\}\right\} \in \mathcal{K}^J$ \label{line:fcycleprolongtrunctwo} \\
    while not \eqref{eq:stoppingcriterion} \\+
        \pr{fascd-vcycle}(J,\ell^J,\underline{\gamma}^J,\overline{\gamma}^J;w^J) \\-
    return $w^J$
\end{pseudo}
\caption{The FAScd F-cycle for solving VI problem \eqref{eq:fe:vi}.}
\label{alg:fascd-fcycle}
\end{pseudofloat}

@@

Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} were implemented in Python, for $P_1$ and $Q_1$ elements, using the Firedrake library \cite{Rathgeberetal2016}.  Here we show numerical results for the example problems presented in Section \ref{sec:vi}, with a focus on the scaling of FAScd iterations as the mesh resolution increases.  Our primary observations are that each V-cycle greatly reduces the residual norm, and that very few F-cycles solve most examples to within discretization error.

@@

Table \ref{tab:results:classical} shows the resulting iteration counts.  When the RS Newton smoother is applied by itself as a solver (``RS only''), as expected it shows rapidly-increasing iterations, roughly proportional to resolution.  By contrast, V-cycle iterations increase slowly, perhaps logarithmically with the resolution.  For F-cycles we report the number of V-cycles after the ramp (Section \ref{sec:implementation}); a constant number solves the problem.  Comparing the ball and spiral results, the complexity of the coincidence set or free boundary does not appear to strongly influence FAScd performance.  For the spiral results, note that \cite{GraeserKornhuber2009} reports more than 30 iterations of all their tested methods were required to achieve similar tolerance for $m_J \approx 5\times 10^5$.

@@

% see git@bitbucket.org:pefarrell/fascd.git: examples/drape.py
% and https://github.com/bueler/mcd-extended: paper/results/drape.sh|txt
\begin{table}[ht]
\begin{tabular}{lr@{\hskip 7mm}cccccc}
\multirow{2}{*}{\emph{levels}} & \multirow{2}{*}{$m_J$} & \multicolumn{3}{c}{\,\emph{ball}} & \multicolumn{3}{c}{\,\emph{spiral}} \\ \cmidrule(lr){3-5} \cmidrule(lr){6-8}
   &          & V-cycle & F-cycle & RS only & V-cycle & F-cycle & RS only \\ \cmidrule{1-8}
 1 &    $5^2$ &  1 &  1 &  1  &  1 &  1 &  1 \\
 2 &    $9^2$ &  2 &  1 &  2  &  2 &  1 &  2 \\
 3 &   $17^2$ &  2 &  2 &  3  &  3 &  2 &  4 \\
 4 &   $33^2$ &  3 &  2 &  6  &  3 &  2 &  7 \\
 5 &   $65^2$ &  3 &  2 &  8  &  4 &  3 &  9 \\
 6 &  $129^2$ &  4 &  2 & 17  &  4 &  3 & 16 \\
 7 &  $257^2$ &  4 &  2 & 33  &  4 &  3 & 27 \\
 8 &  $513^2$ &  5 &  2 & 65  &  5 &  3 & 48 \\
 9 & $1025^2$ &  5 &  2 & \XX &  5 &  3 & \XX \\
10 & $2049^2$ &  5 &  2 & \XX &  5 &  3 & \XX
\end{tabular}
\bigskip
\caption{Iterations for FAScd V-cycle and F-cycle algorithms in Example \ref{ex:results:classical}.  Smoother-only runs marked \XX\xspace were not attempted.}
\label{tab:results:classical}
\end{table}

@@

% figures in the following visualization were generated using branch elbueler/simplefigs
% of git@bitbucket.org:pefarrell/fascd.git, and the following commands (including imagemagick):
%   cd examples/
%   python3 plap1d.py -fascd_cycle_type multiplicative -mcoarse 4 -levs 3 -figures
%   [save the images into mcd-extended/paper/fixfigs/vcycle/]
%   for X in start2 down2 down1 up0 up1 up2 next2; do mogrify -trim $X.png $X.png; done
\begin{figure}[ht]
\begin{center}
\input{tikz/imagesvcycle.tex}
\end{center}
\caption{Visualization of one FAScd V-cycle iteration for Example \ref{ex:results:plap}, with $p=1.5$, $J=3$, and a coarsest mesh of $m_0=5$ vertices.  Solutions (i.e.~$w^j$, $y^j$, $z^j$) are lines, obstacles ($\underline{\gamma}^j$, $\underline{\phi}^j$, $\underline{\chi}^j$) are strong dots, and the zero function is dotted.  Vertical scale varies.}
\label{fig:imagesvcycle}
\end{figure}

@@

Solvers implemented in Firedrake can exploit parallel solvers from the PETSc library \cite{Balayetal2023}, and in fact the above Examples, which were run in serial so as to fix the smoother, also run efficiently in parallel.  The next Examples apply parallel solvers, and Example \ref{ex:results:sia} demonstrates the parallel weak scaling \cite{Bueler2021} of FAScd.

@@

F-cycle iteration results are shown in Table \ref{tab:results:advdiff}.  Despite the hint of growing iterations under mesh refinement, a few iterations sufficed up to memory-limited resolutions.  For any of the higher resolution runs, the FAScd F-cycle solver took less time than generating the $P_1$ mesh hierarchy.  The local Péclet number $\text{Pe}=\|\bX(x,y,z)\|_2 L \eps^{-1}$, with $L=2$ the width of $\Omega$, varied from 40 to 260; the mesh Péclet number $m_J^{-1/d}\,\text{Pe}$ varied accordingly.  Thus the $m_J\le 121^d$ problems were advection-dominated while those with $m_J\ge 481^d$ were diffusion-dominated, but the smoother was the same in all cases, and the iterations were nearly constant.
\end{example}

@@

In contrast with monotone multigrid \cite{Kornhuber1994}, FAScd does not modify the FE discretization, but the possibility of acceleration via basis-level manipulations is a topic for future research.  On the other hand,  the convergence proofs in \cite{GraeserKornhuber2009,Kornhuber1994}, which appear to depend on particular smoother and basis-level manipulation choices, have no analog here.

>>>>

\begin{abstract}
We present the \emph{full approximation scheme constraint decomposition} (FASCD) multilevel method for solving variational inequalities (VIs).  FASCD is a common extension both of the full approximation scheme (FAS) multigrid technique for nonlinear partial differential equations, due to A.~Brandt, and the constraint decomposition (CD) method introduced by X.-C.~Tai for VIs arising in optimization.  We extend the CD idea by exploiting the telescoping nature of certain function space subset decompositions arising from multilevel mesh hierarchies.  We demonstrate that FASCD V-cycles exhibit excellent mesh-independent convergence rates and that FASCD F-cycles (full multigrid) are optimal solvers, when used in combination with reduced-space active set smoothers with work proportional to the number of unknowns on a given mesh level.  The example problems include unilateral and bilateral VI problems, with differential operators which are symmetric linear, nonsymmetric linear, and nonlinear.  Our examples demonstrate very good convergence even on problems where other multilevel VI solvers struggle.
\end{abstract}

@@

The constraint decomposition (CD) methods of Tai \cite{Tai2003} are designed to solve coercive variational inequality (VI) problems arising as the Karush--Kuhn--Tucker optimality conditions for the minimization of a convex functional over a convex set.  These methods were shown to converge for multilevel and domain decompositions in a finite element (FE) context; in the multilevel case, the CD method has almost grid-independent error bounds, and near-optimal complexity, for elliptic, linear obstacle problems \cite[Subsection 5.4]{Tai2003}, \cite[Theorem 4.6 and Algorithm 4.7]{GraeserKornhuber2009}.
In this paper we extend Tai's multilevel CD method by replacing the coarser-level corrections with a full approximation scheme (FAS; \cite{Brandt1977,Bruneetal2015}) approach.  Specifically, our \emph{full approximation scheme constraint decomposition} (FASCD) method extends the algorithms in \cite{GraeserKornhuber2009} and \cite{Tai2003} in the following five directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item The FAS coarse corrections are adapted for general VI problems. The needed computations require neither an objective function nor linearity of the residual operator.
\item We identify a ``telescoping'' subset property which can be exploited on the up-smoothing side of a V-cycle for greater performance.
\item Box constraints, namely upper and lower obstacles, are permitted.
\item We implement V-cycles (Algorithm \ref{alg:fascd}) and F-cycles (full multigrid; Algorithm \ref{alg:fascd-fcycle}).
\item We demonstrate that F-cycle FASCD is an optimal solver for the different VI examples in Section \ref{sec:results}, when used in conjunction with reduced-space active set methods as smoothers. The number of Newton and Krylov iterations per step are fixed, so that the work of the smoother is proportional to the number of unknowns on a given level.
\end{enumerate}

Prior multilevel algorithms for VIs have various substantial limitations compared to the FASCD framework.  The projected full approximation scheme (PFAS) algorithm of Brandt \& Cryer \cite{BrandtCryer1983} is apparently limited to linear complementarity problems and projected (truncated) Gauss--Seidel smoothers.  While FASCD reduces to general, nonlinear FAS for PDEs when the inequality constraints are removed (Appendix \ref{app:reductions}), PFAS reduces to linear multigrid.
The multilevel CD iteration of Tai \cite{Tai2003} is formulated in terms of an objective function, so as stated it only applies to optimization-type problems, and furthermore each of its coarser-level problems refers to the finest-level discretization.  The problematic latter property is corrected by reformulation using defect constraints and monotone injection operators (see \cite[Algorithm 4.7]{GraeserKornhuber2009} and Section \ref{sec:cdmultilevel} herein).  The monotone multigrid methods of Kornhuber and co-authors \cite{GraeserKornhuber2009,Kornhuber1994} appear to be formulated only with Gauss--Seidel type smoothers, and applied only to optimization-type VI problems \cite[for example]{JouvetGraeser2013}.  Furthermore, monotone multigrid methods are more difficult to implement in FE libraries because modifications of the FE basis functions are needed to approximate the coarser-level coincidence sets.  No such invasive basis modifications are used either in Tai's original multilevel CD or in FASCD.  Our implementation of FASCD in the Firedrake library \cite{Rathgeberetal2016} requires no modifications of basis functions and yet exhibits excellent performance, including on a problem used to demonstrate the benefits of monotone multigrid (Example \ref{ex:results:classical}; compare \cite[problem 7.1.1]{GraeserKornhuber2009}).

In our experiments, FASCD V-cycles exhibit essentially the same mesh-independent convergence rates for VI problems that FAS multigrid V-cycles exhibit for the corresponding (unconstrained) PDE problems.  The F-cycle results in Section \ref{sec:results} show nearly-textbook multigrid efficiency, namely solutions within discretization error in work comparable to few smoother sweeps on the finest-level mesh \cite{BrandtLivne2011}.  Specifically, we see this performance for classical Laplacian obstacle problems, $p$-Laplacian obstacle problems, and nonnegativity-constrained advection-diffusion equations.  Furthermore, these results appear to be insensitive to the length or structure of the free boundary, i.e.~the geometric complexity of the coincidence set.

All iterates in FASCD are admissible, and thus the nonlinear operator in the VI needs only be defined for admissible states.  For example, we solve an ice sheet geometry problem (Examples \ref{ex:sia} and \ref{ex:results:sia}) in which the ice flow formulas are only meaningful for surface elevation iterates that do not penetrate the bedrock.  (More generally, fluid layers subject to a non-negative thickness constraint typically require the solution of an auxiliary PDE on a domain determined inside the VI residual evaluation \cite{Bueler2021conservation}, but this is not needed in the ice sheet problem here.)  Non-admissible methods, including semi-smooth methods \cite{BensonMunson2006}, would require unnatural modifications of the operator formula.  FASCD solves this low-regularity ice sheet problem, wherein the solution has unbounded gradient at the free boundary, by a mesh-independent number of F-cycle iterations (Example \ref{ex:results:sia}).

The paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone variational inequalities, and describe several motivating example problems.  Section \ref{sec:cd} reviews the basics of CDs and their associated iterations.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, and then Section \ref{sec:cdmultilevel} presents defect constraints generated via monotone injection operators \cite{GraeserKornhuber2009}.  Here we propose an apparently new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD and an associated iteration.  Section \ref{sec:vcycle} then states the FASCD V-cycle, with Section \ref{sec:implementation} addressing convergence criteria, F-cycles, and smoother construction.  Section \ref{sec:results} gives numerical results, from a parallel Firedrake \cite{Rathgeberetal2016} implementation (see Code Availability), for each example problem introduced in Section \ref{sec:vi}.  The work concludes with a discussion and outlook in Section \ref{sec:discussion}.

@@

Numerical results are shown in Section \ref{sec:results} for applications of FASCD Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} to these Examples.

@@

Associated to any CD are certain iterative methods \cite{Tai2003,Xu1992}.  Let $0\le \alpha \le 1$ be the \emph{damping parameter}.\footnote{FASCD Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} are presented without damping, and indeed it seems to be unnecessary.  However, the convergence of the general CD iterations appears to depend upon it.}  The following algorithms, which replace $u \in \cK$ with a new iterate $w\in\cK$ which should be closer to $u^\star \in \cK$ solving \eqref{eq:vi}, solve smaller VI problems over each subset $\cK_i$.

@@

\section{FASCD V-cycle} \label{sec:vcycle}

@@

\begin{figure}[ht]
\begin{center}
\input{tikz/vcycle.tex}
\end{center}
\caption{The FASCD V-cycle computes downward corrections $y_j \in \mathcal{D}^j$, but the upward corrections $z_j\in\mathcal{U}^j$ are in larger sets.}
\label{fig:fascdvcycle}
\end{figure}

@@

\begin{pseudofloat}[ht]
\begin{pseudo}
\pr{fascd-vcycle}(J,\ell^J,\underline{\gamma}^J,\overline{\gamma}^J;w^J)\text{:} \\+
    $\underline{\chi}^J, \, \overline{\chi}^J = \underline{\gamma}^J - w^J, \, \overline{\gamma}^J - w^J {\large \strut}$ \label{line:vcyclegenchifinest} \\
    for $j=J$ downto $j=1$ \\+
      $\underline{\chi}^{j-1}, \, \overline{\chi}^{j-1} = \maxR \underline{\chi}^j, \, \minR \overline{\chi}^j {\large \strut}$ \label{line:vcyclegenchi} \\
      $\underline{\phi}^j, \, \overline{\phi}^j = \underline{\chi}^j - P\underline{\chi}^{j-1}, \, \overline{\chi}^j - P\overline{\chi}^{j-1} {\large \strut}$ \label{line:vcyclegenphi} \\
      $y^j = 0$ \\
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j,\underline{\phi}^j,\overline{\phi}^j,w^j;y^j)$  \ct{smoothing of \eqref{eq:fe:downvi} in $\mathcal{D}^j$}\\
      $w^{j-1} = \iR(w^j + y^j)$ \label{line:vcyclerestrictsolution} \\
      $\ell^{j-1} = f^{j-1}(w^{j-1}) + R \left(\ell^j - f^j(w^j+y^j)\right)$ \label{line:vcyclerestrictell} \\-
    $z^0 = 0$ \\
    $\text{\pr{solve}}(\ell^0,\underline{\chi}^0,\overline{\chi}^0,w^0;z^0)$ \hspace{1.0cm} \ct{solving of \eqref{eq:fe:upvi} in $\mathcal{U}^0$} \\
    for $j=1$ to $j=J$ \\+
      $z^j = y^{j} + P z^{j-1}$ \label{line:vcycleupsmoothinitial} \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j,\underline{\chi}^j,\overline{\chi}^j,w^j;z^j)$  \ct{smoothing of \eqref{eq:fe:upvi} in $\mathcal{U}^j$} \\-
    return $w^J+z^J$
\end{pseudo}
\caption{The FASCD V-cycle is an iteration for solving VI problem \eqref{eq:fe:vi}.  Here $f^j$ denotes an FE discretization of $f$ in problem \eqref{eq:boxdirichletvi}.}
\label{alg:fascd}
\end{pseudofloat}

@@

Implementations of FASCD methods must choose a concrete convergence criterion.  In this section we also construct the F-cycle extension, and we describe the active set Newton method we will use as a smoother and coarse solver.

@@

\begin{pseudofloat}[h]
\begin{pseudo}
\pr{fascd-fcycle}(J,\ell^J,\underline{\gamma}^J,\overline{\gamma}^J)\text{:} \\+
    for $j=J$ downto $j=1$ \\+
        $\ell^{j-1} = R(\ell^j)$ \\
        $\underline{\gamma}^{j-1}, \, \overline{\gamma}^{j-1} = \iR \underline{\gamma}^{j}, \, \iR \overline{\gamma}^{j}$ {\large\strut} \\-
    choose $w^0 \in \mathcal{K}^0$ \label{line:fcyclecoarsestinitial} \\
    $\text{\pr{solve}}(\ell^0,\underline{\gamma}^0,\overline{\gamma}^0;w^0)$ \\
    for $j=1$ to $j=J-1$ \\+
        $w^j = \max\left\{\underline{\gamma}^{j},\min\{\overline{\gamma}^{j}, Pw^{j-1}\}\right\} \in \mathcal{K}^j$ \label{line:fcycleprolongtruncone} \\
        $\pr{fascd-vcycle}^{\text{\id{rampv}}}(j,\ell^j,\underline{\gamma}^j,\overline{\gamma}^j;w^j)$ \\-
    $w^J = \max\left\{\underline{\gamma}^{J},\min\{\overline{\gamma}^{J}, Pw^{J-1}\}\right\} \in \mathcal{K}^J$ \label{line:fcycleprolongtrunctwo} \\
    while not \eqref{eq:stoppingcriterion} \\+
        \pr{fascd-vcycle}(J,\ell^J,\underline{\gamma}^J,\overline{\gamma}^J;w^J) \\-
    return $w^J$
\end{pseudo}
\caption{The FASCD F-cycle for solving VI problem \eqref{eq:fe:vi}.}
\label{alg:fascd-fcycle}
\end{pseudofloat}

@@

Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} were implemented in Python, for $P_1$ and $Q_1$ elements, using the Firedrake library \cite{Rathgeberetal2016}.  Here we show numerical results for the example problems presented in Section \ref{sec:vi}, with a focus on the scaling of FASCD iterations as the mesh resolution increases.  Our primary observations are that each V-cycle greatly reduces the residual norm, and that very few F-cycles solve most examples to within discretization error.

@@

Table \ref{tab:results:classical} shows the resulting iteration counts.  When the RS Newton smoother is applied by itself as a solver (``RS only''), as expected it shows rapidly-increasing iterations, roughly proportional to resolution.  By contrast, V-cycle iterations increase slowly, perhaps logarithmically with the resolution.  For F-cycles we report the number of V-cycles after the ramp (Section \ref{sec:implementation}); a constant number solves the problem.  Comparing the ball and spiral results, the complexity of the coincidence set or free boundary does not appear to strongly influence FASCD performance.  For the spiral results, note that \cite{GraeserKornhuber2009} reports more than 30 iterations of all their tested methods were required to achieve similar tolerance for $m_J \approx 5\times 10^5$.

@@

% see git@bitbucket.org:pefarrell/fascd.git: examples/drape.py
% and https://github.com/bueler/mcd-extended: paper/results/drape.sh|txt
\begin{table}[ht]
\begin{tabular}{lr@{\hskip 7mm}cccccc}
\multirow{2}{*}{\emph{levels}} & \multirow{2}{*}{$m_J$} & \multicolumn{3}{c}{\,\emph{ball}} & \multicolumn{3}{c}{\,\emph{spiral}} \\ \cmidrule(lr){3-5} \cmidrule(lr){6-8}
   &          & V-cycle & F-cycle & RS only & V-cycle & F-cycle & RS only \\ \cmidrule{1-8}
 1 &    $5^2$ &  1 &  1 &  1  &  1 &  1 &  1 \\
 2 &    $9^2$ &  2 &  1 &  2  &  2 &  1 &  2 \\
 3 &   $17^2$ &  2 &  2 &  3  &  3 &  2 &  4 \\
 4 &   $33^2$ &  3 &  2 &  6  &  3 &  2 &  7 \\
 5 &   $65^2$ &  3 &  2 &  8  &  4 &  3 &  9 \\
 6 &  $129^2$ &  4 &  2 & 17  &  4 &  3 & 16 \\
 7 &  $257^2$ &  4 &  2 & 33  &  4 &  3 & 27 \\
 8 &  $513^2$ &  5 &  2 & 65  &  5 &  3 & 48 \\
 9 & $1025^2$ &  5 &  2 & \XX &  5 &  3 & \XX \\
10 & $2049^2$ &  5 &  2 & \XX &  5 &  3 & \XX
\end{tabular}
\bigskip
\caption{Iterations for FASCD V-cycle and F-cycle algorithms in Example \ref{ex:results:classical}.  Smoother-only runs marked \XX\xspace were not attempted.}
\label{tab:results:classical}
\end{table}

@@

% figures in the following visualization were generated using branch elbueler/simplefigs
% of git@bitbucket.org:pefarrell/fascd.git, and the following commands (including imagemagick):
%   cd examples/
%   python3 plap1d.py -fascd_cycle_type multiplicative -mcoarse 4 -levs 3 -figures
%   [save the images into mcd-extended/paper/fixfigs/vcycle/]
%   for X in start2 down2 down1 up0 up1 up2 next2; do mogrify -trim $X.png $X.png; done
\begin{figure}[ht]
\begin{center}
\input{tikz/imagesvcycle.tex}
\end{center}
\caption{Visualization of one FASCD V-cycle iteration for Example \ref{ex:results:plap}, with $p=1.5$, $J=3$, and a coarsest mesh of $m_0=5$ vertices.  Solutions (i.e.~$w^j$, $y^j$, $z^j$) are lines, obstacles ($\underline{\gamma}^j$, $\underline{\phi}^j$, $\underline{\chi}^j$) are strong dots, and the zero function is dotted.  Vertical scale varies.}
\label{fig:imagesvcycle}
\end{figure}

@@

Solvers implemented in Firedrake can exploit parallel solvers from the PETSc library \cite{Balayetal2023}, and in fact the above Examples, which were run in serial so as to fix the smoother, also run efficiently in parallel.  The next Examples apply parallel solvers, and Example \ref{ex:results:sia} demonstrates the parallel weak scaling \cite{Bueler2021} of FASCD.

@@

F-cycle iteration results are shown in Table \ref{tab:results:advdiff}.  Despite the hint of growing iterations under mesh refinement, a few iterations sufficed up to memory-limited resolutions.  For any of the higher resolution runs, the FASCD F-cycle solver took less time than generating the $P_1$ mesh hierarchy.  The local Péclet number $\text{Pe}=\|\bX(x,y,z)\|_2 L \eps^{-1}$, with $L=2$ the width of $\Omega$, varied from 40 to 260; the mesh Péclet number $m_J^{-1/d}\,\text{Pe}$ varied accordingly.  Thus the $m_J\le 121^d$ problems were advection-dominated while those with $m_J\ge 481^d$ were diffusion-dominated, but the smoother was the same in all cases, and the iterations were nearly constant.
\end{example}

@@

In contrast with monotone multigrid \cite{Kornhuber1994}, FASCD does not modify the FE discretization, but the possibility of acceleration via basis-level manipulations is a topic for future research.  On the other hand,  the convergence proofs in \cite{GraeserKornhuber2009,Kornhuber1994}, which appear to depend on particular smoother and basis-level manipulation choices, have no analog here.
